\documentclass{article}
\usepackage{amsmath,amsthm,amssymb}

\newenvironment{problem}[2][Problem]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}

\begin{document}

\title{Problem Set 1, Answers}
\author{Kevin Lacker}
\maketitle

\begin{problem}{1}

  A language $L$ is in $\Sigma_2^{\bf{P}}$ iff there is a polynomial time TM
  $M$ such that:

  \begin{equation}
    x \in L \iff \exists u_1 \forall u_2 M(x, u_1, u_2)
  \end{equation}

  where the $u_i$ are polynomial size and we treat $M$ as returning
  true or false. This is equivalent to:

  \begin{equation}
    x \in L \iff \exists u_1 \neg ( \exists u_2 ( \neg M(x, u_1, u_2) )
  \end{equation}

  The answer to $\exists u_2 ( \neg M(x, u_1, u_2) )$ can be found
  in a single call to an ${\bf{NP}}$ oracle, so $\Sigma_2^{\bf{P}} \in
  \bf{NP}^{\bf{NP}}$ .

  The other direction is similar. For any language $L$ in
  $\bf{NP}^{\bf{NP}}$, there is a Turing machine $M$, where $M$ has
  access to an ${\bf{NP}}$ oracle, such that:

  \begin{equation}
    x \in L \iff \exists v_1 M(x, v_1)
  \end{equation}

  There are different equivalent formats for the oracle. It is
  convenient to think of the input as being a $SAT$ problem, and the
  output is either a solution or an indication that there is no
  solution.

  To write this as a $\Sigma_2^{\bf{P}}$ language we need to convert
  the oracle calls to a regular Turing machine under a $\forall$
  quantifier. So, let $v_2$ represent the oracle outputs, and $v_3$
  represent a string of
  bits long enough to hold all assignments to the $SAT$ problems where
  the oracle reported no solution. $M$ makes polynomially many calls
  to the oracle, so the lenght of $v_2$ and $v_3$ can be polynomially
  bounded.

  Then, we can represent $L$ as:

  \begin{equation}
    x \in L \iff \exists v_1 v_2 \forall v_3 M'(x, v_1, v_2, v_3)
  \end{equation}

  Where $M'$ works like:
  
  \begin{itemize}
    \item It performs the same operations as $M$, except when there is
      an oracle call
    \item When the oracle gives a solution, $M'$ validates the
      solution
    \item When the oracle reports no solution, $M'$ uses bits from
      $v_3$ to check that the particular inputs do not represent a solution
  \end{itemize}

  Since this is quantified over all $v_3$, the quantifiers do the job
  of the $NP$ oracle, and this is an equivalent representation for
  $L$. Thus $\bf{NP}^{\bf{NP}} \in \Sigma_2^{\bf{P}}$.
  
\end{problem}

\begin{problem}{2}

  Assume for the sake of contradiction that $NP = SPACE(n)$. For any
  $L \in SPACE(n^2)$, we can pad the length to $n^2$ to get a
  language that is in $SPACE(n)$. By our assumption, this language is
  in $NP$.

  However, if this padded language is in $NP$, it must also be in $NP$
  without the padding, since a nondeterministic machine can add the
  padding, and runtime polynomial in $n^2$ is also polynomial in
  $n$. So $L$ is in $NP$ as well. But since $NP = SPACE(n)$, this
  shows that $SPACE(n^2) \subset SPACE(n)$, which contradicts the
  space hierarchy theorem.
  
\end{problem}

\begin{problem}{3}
  TODO
\end{problem}

\begin{problem}{4}
  Assume for the sake of contradiction that $\Sigma_2^P \in
  SIZE[n^k]$. Since $NP \in \Sigma_2^P$, this implies $NP \in P/poly$,
  and so by the Karp-Lipton theorem, the polynomial hierarchy
  collapses, with $\Sigma_2^P = \Sigma_3^P = PH$. We can then
  substitute into our assumption to get $\Sigma_3^P \in SIZE[n^k]$.

  However, $\Sigma_3^P$ cannot have polynomial circuits, because it
  has enough power to find a circuit with a given complexity. It will
  be useful to have a total ordering of all circuits, first by number
  of gates, with tiebreaks broken lexicographically. So a ``smaller''
  circuit can either be one with fewer gates, or one earlier
  lexicographically.
  Then consider this chain of problems:

  \begin{itemize}
    \item Given two circuits, is there any input where their result
      differs? This problem is in $NP$.
    \item Given a circuit, is there any smaller circuit
      that computes the same result on all inputs? This can be
      solved in $NP$ with an oracle for the previous problem, so it
      is in $NP^{NP} = \Sigma_2^P$.
    \item Find the smallest circuit of size $n^{k+1}$ that has no
      smaller circuit that computes the same result on all
      inputs. (This exists, by a counting argument.) This can be
      solved in $NP$ with an oracle for the previous problem, so it
      is in $NP^{{NP}^{NP}} = \Sigma_3^P$.
  \end{itemize}

  A function that computes the result of the circuit output by the
  last algorithm has no circuits smaller than $n^{k+1}$, but it is in
  $\Sigma_3^P$, which leads us to a contradiction.
\end{problem}

\begin{problem}{5}
  Assume we have $P = NP$ and consider the $NEXP$-complete problem of
  $SUCCINCT$-$SAT$. In exponential time, we can expand out the
  truth table of the compressed $SAT$ problem into an exponentially
  long $SAT$ problem. We can then use our polynomial-time algorithm
  for solving $NP$ problems on this exponentially long $SAT$
  problem. It takes time that is the polynomial of an exponential, but
  this is still contained in $EXP$. So $P = NP$ implies $EXP = NEXP$.

  In particular, when $P = NP$ you can find a circuit that requires at
  least $x$ gates with time polynomial in $x$. By the reasoning in the
  previous problem, this task is in $\Sigma_3^P$, and when $P = NP$
  the hierarchy collapses and this task is in $P$.

  We know from Shannon's counting argument that there exists some function on
  $n$ bits that requires $2^n / n$ gates. So we just have to find
  it. We can use this algorithm, and it takes time polynomial in $2^n / n$.
  A polynomial function of $2^n$ is bounded by $2^{cn}$ for some $c$, and
  these algorithms are contained in $EXP$. So if $P = NP$, the 
  language determined by the first such circuit is in $EXP$.

  (This didn't use $EXP = NEXP$ directly, as hinted, but I think the reasoning is
  simpler this way.)
\end{problem}

\begin{problem}{6}
  So we have a polynomial-time algorithm that can use a single query to an $NP$
  oracle to determine the satisfiability to two $3SAT$ formulas
  $\Phi_1$ and $\Phi_2$. We can remove the oracle and run the
  algorithm twice. The first time, we pretend the oracle returns a
  ``true''. The second time, we pretend the oracle returns a
  ``false''. Each time our algorithm gives us a set of two answers -
  whether $\Phi_1$ is satisfiable, and whether $\Phi_2$ is
  satisfiable. And we know one of these two sets is correct, although
  we don't have the oracle so we don't know which set it is. You can
  think of this as two ``branches'' of the algorithm. We get an answer
  from each branch, and we don't know which branch is the correct one.

  So each branch can report one of four things:

  \begin{itemize}
  \item $\Phi_0$ and $\Phi_1$ are both satisfiable
  \item neither of $\Phi_0$ and $\Phi_1$ are satisfiable
  \item $\Phi_0$ is satisfiable, $\Phi_1$ is not
  \item $\Phi_1$ is satisfiable, $\Phi_0$ is not    
  \end{itemize}

  What will we learn from these results? If both branches report the
  same thing about either formula, this tells us the answer for that formula. If
  the branches do not agree on either formula, there are two
  cases to consider. One branch could be saying ``both satisfiable''
  and the other could be saying ``neither''. In this case, we learn
  that the formulas have the same satisfiability. Or, the branches
  could both be saying different formulas are the satisfiable ones. In
  this case, we learn that the formulas have opposite
  satisfiability. In other words, we have a polynomial-time algorithm
  that tells us one of the following things:

  \begin{itemize}
    \item $\Phi_0$ is satisfiable
    \item $\Phi_0$ is not satisfiable
    \item $\Phi_1$ is satisfiable
    \item $\Phi_1$ is not satisfiable
    \item Either both or neither of the $\Phi_i$ are satisfiable
    \item Precisely one of the $\Phi_i$ are satisfiable
  \end{itemize}

  We can now use this recursively to solve $3SAT$ problems. For a
  formula $\Phi$, create two smaller formulas. Substitute $x_0 = 0$ to
  get $\Phi_0$, and substitute $x_0 = 1$ to get $\Phi_1$. Now, $\Phi$
  is satisfiable if either $\Phi_0$ or $\Phi_1$ is satisfiable.

  We run our algorithm on $\Phi_0$ and $\Phi_1$ and the way we recurse depends on
  the information we get.

  \begin{itemize}
    \item If either $\Phi_i$ is satisfiable, $\Phi$ is satisfiable and
      we are done.
    \item If either $\Phi_i$ is not satisfiable, we can recurse on the
      other one.
    \item If the $\Phi_i$ are either both or neither satisfiable, we
      can recurse on either one.
    \item If precisely one is satisfiable, then $\Phi$ is
      satisfiable and we are done.
  \end{itemize}

  Each time we recurse it takes polynomial time and we remove one free
  variable, so the whole algorithm will take polynomial time, and we
  have a polynomial time algorithm for solving $3SAT$, so $P = NP$.
\end{problem}

\end{document}
